---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit only README.Rmd! -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

```{r, echo = FALSE}
library(DNAtools)
```

# DNAtools

[![Build Status](https://travis-ci.org/mikldk/DNAtools.svg?branch=master)](https://travis-ci.org/mikldk/DNAtools)
[![Build status](https://ci.appveyor.com/api/projects/status/1861od7todeskm5p/branch/master?svg=true)](https://ci.appveyor.com/project/mikldk/DNAtools/branch/master)

Development version of the DNAtools R-package.

## Install

To build and install from Github using R 3.3.0 (or later) and the R `devtools` package 1.11.0 (or later) run this command from within `R`:

```
devtools::install_github("mikldk/DNAtools", 
                         build_opts = c("--no-resave-data", "--no-manual"))
```

You can also install the package without vignettes if needed as follows:

```
devtools::install_github("mikldk/DNAtools")
```


## Contribute, issues, and support ##

Please use the issue tracker at <https://github.com/mikldk/DNAtools/issues> 
if you want to notify us of an issue or need support.
If you want to contribute, please either create an issue or make a pull request.

## Getting started

See documentation included in package (vignettes and manual) at <https://mikldk.github.io/DNAtools/>.

### Small example

Please read the vignettes for more elaborate explanations than those given below. 
The below example is meant to illustrate some of the functionality the package provides in 
a compact fashion.

Say that we have a reference database:

```{r}
data(dbExample, package = "DNAtools")
head(dbExample)[, 2:7]
dim(dbExample)
```

One could ask: What is the distribution of the number of alleles observed in a three person mixture?
This can be easily calculated by this package. First we find the allele frequencies:

```{r}
allele_freqs <- lapply(1:10, function(x){
  al_freq <- table(c(dbExample[[x*2]], dbExample[[1+x*2]]))/(2*nrow(dbExample))
  al_freq[sort.list(as.numeric(names(al_freq)))]
})
names(allele_freqs) <- sub("\\.1", "", names(dbExample)[(1:10)*2])
```

Now, the distribution of the number of alleles in a three person mixture can be calculated. 
We focus on the D16S539 locus:

```{r}
allele_freqs$D16S539
noa <- Pnm_locus(m = 3, theta = 0, alleleProbs = allele_freqs$D16S539)
names(noa) <- seq_along(noa)
noa
```

So it is most likely that a three person mixture on D16S539 has `r names(noa)[which.max(noa)]` alleles.

This can be done for all loci at once:

```{r}
noa <- Pnm_all(m = 3, theta = 0, probs = allele_freqs, locuswise = TRUE)
noa
```

We can also find the convolution and thereby the total number of distinct alleles:

```{r}
noa <- Pnm_all(m = 3, theta = 0, probs = allele_freqs)
noa
```

So it is most likely that a three person mixture has `r names(noa)[which.max(noa)]` distinct alleles on all loci combined.



