---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit only README.Rmd! -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

```{r, echo = FALSE}
library(DNAtools)
```

# DNAtools

[![Build Status](https://travis-ci.org/mikldk/DNAtools.svg?branch=master)](https://travis-ci.org/mikldk/DNAtools)
[![Build status](https://ci.appveyor.com/api/projects/status/1861od7todeskm5p/branch/master?svg=true)](https://ci.appveyor.com/project/mikldk/DNAtools/branch/master)
?[![DOI](https://joss.theoj.org/papers/10.21105/joss.01981/status.svg)](https://doi.org/10.21105/joss.01981)

There are two main features of this package:

* Computation of the distribution of the numbers of alleles in DNA mixtures.
* Empirical testing of DNA match probabilities.

Each is described in a separate vignette, and a small example given 
below under "Getting started". 
The documentation (vignettes and manual) is both included in package 
and available for reading online at <https://mikldk.github.io/DNAtools/>.


## Install

### With internet access

To build and install from Github using R 3.3.0 (or later) and the R `devtools` package 1.11.0 (or later) run this command from within `R`:

```
devtools::install_github("mikldk/DNAtools", 
                         build_opts = c("--no-resave-data", "--no-manual"))
```

You can also install the package without vignettes if needed as follows:

```
devtools::install_github("mikldk/DNAtools")
```

### Without internet access

To install on a computer without internet access:

1. Download `DNAtools` as a `.tar.gz` archive from GitHub, transfer to the destination computer, e.g. using removable media
1. Install `devtools` and `DNAtools` pre-requisites (`multicool`, `Rcpp`, `RcppParallel`, `RcppProgress`, `Rsolnp`)
1. Install `DNAtools` in `R` using the `devtools::install_local()` function


## Contribute, issues, and support ##

Please use the issue tracker at <https://github.com/mikldk/DNAtools/issues> 
if you want to notify us of an issue or need support.
If you want to contribute, please either create an issue or make a pull request.

## Getting started


Please read the vignettes for more elaborate explanations than those given below. 
The below example is meant to illustrate some of the functionality the package provides in 
a compact fashion.

Say that we have a reference database:

```{r}
data(dbExample, package = "DNAtools")
head(dbExample)[, 2:7]
dim(dbExample)
```

We now find the allele frequencies:

```{r}
allele_freqs <- lapply(1:10, function(x){
  al_freq <- table(c(dbExample[[x*2]], dbExample[[1+x*2]]))/(2*nrow(dbExample))
  al_freq[sort.list(as.numeric(names(al_freq)))]
})
names(allele_freqs) <- sub("\\.1", "", names(dbExample)[(1:10)*2])
```


```{r, include=FALSE}
txtbar <- function(x) {
  y <- round(100*noa)
  y2 <- lapply(y, rep.int, x = "|")
  y3 <- lapply(y2, paste0, collapse = "")
  ret <- data.frame(`Number of alleles` = names(x), Frequency = unlist(y3), 
                    check.names = FALSE)
  print(ret, quote = FALSE, row.names = FALSE, right = FALSE)
  return(invisible(ret))
}
```

### Number of alleles

One could ask: What is the distribution of the number of alleles observed in a three person mixture?

The distribution of the number of alleles in a three person mixture can 
be calculated by this package. 
We focus on the D16S539 locus:

```{r}
allele_freqs$D16S539
noa <- Pnm_locus(m = 3, theta = 0, alleleProbs = allele_freqs$D16S539)
names(noa) <- seq_along(noa)
noa
```

This can be illustrated by a barchart:

```{r, echo=FALSE, results='markup', comment=''}
txtbar(noa)
```

So it is most likely that a three person mixture on D16S539 has `r names(noa)[which.max(noa)]` alleles.

This can be done for all loci at once:

```{r}
noa <- Pnm_all(m = 3, theta = 0, probs = allele_freqs, locuswise = TRUE)
noa
```

We can also find the convolution and thereby the total number of distinct alleles:

```{r}
noa <- Pnm_all(m = 3, theta = 0, probs = allele_freqs)
noa
```

This can be illustrated by a barchart:

```{r, echo=FALSE, results='markup', comment=''}
txtbar(noa)
```

So it is most likely that a three person mixture has `r names(noa)[which.max(noa)]` distinct alleles on all loci combined.


### Empirical testing of DNA match probabilities

Another relevant questions is how many matches and near-matches there are. 
This can be calculated as follows:

```{r}
db_summary <- dbCompare(dbExample, hit = 6, trace = FALSE)
db_summary
```

The hit argument returns pairs of profiles that fully match at `hit` (here 6) or more loci.

The summary matrix gives the number of pairs mathcing/partially-matching at $(i,j)$ loci. 
For example the row
```
     partial
match     0     1     2     3     4     5     6     7     8     9    10
   5      6    19    44    41    26     5                              
```
means that there are 6+19+44+41+26+5 = 141 pairs of profiles matching exactly at 
5 loci. 
Conditional on those 5 matches, there are 
6 pairs not matching on the remaining 5 loci, 
19 pairs partial matching on 1 locus and not matching on the remaining 4 loci, 
and so on.


