---
title: "On the exact distribution of the numbers of alleles in DNA mixtures"
author: "Mikkel Meyer Andersen, James Michael Curran and Torben Tvedebrink"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
options(knitr.kable.NA = '')
```

# Introduction

When more than one individual contributes biological material to a
forensic stain, the resulting DNA type is termed a DNA mixture. DNA
mixtures occur frequently in forensic genetic casework, and in
recent years much research has been devoted to this subject.

This vignette shows how to compute the exact distribution of the number 
of alleles for any number of profiles and investigated loci using the `DNAtools`
package in R. 
```{r}
library(DNAtools)
```


The per locus number of observed alleles is of interest as it indicates the
plausible range of the number of contributors. If a DNA mixture has $m$ 
contributors, it is possible to observe anything between $1$ and $2m$ alleles
at a given marker. Of particular interest is the probability at which an $m+1$
person DNA mixture is percieved as an $m$ person DNA mixture. That is, what is 
the probability that $m$ persons have at most $2(m-1)$ distinct alleles among
them?

# DNA mixtures

In forensic genetics, a DNA mixture refers to the situation where more
than one individual's DNA is present in an analysed trace. DNA
mixtures are regularly encountered in forensic genetics.

However, an exact expression of the number of alleles has not been
presented up to now. We show how to derive the probability that $N$
alleles are observed given $m$ contributors and $L$ typed loci.

The prevailing technology for identity assessment and DNA mixture analysis in 
forensic genetics uses so-called Short Tandem Repeat (STR) markers or loci.
The number of alleles observed at these markers are typically in the range of
10 to 30 alleles. 

In the `DNAtools` there is a database with $1,000$ simulated DNA profiles on
10 autosomal STR markers, which we will use to compute the allele frequencies.
```{r}
data(dbExample, package = "DNAtools")
knitr::kable(head(dbExample)[,2:7])
```

We extract the relevant columns and compute the allele frequencies by the 
allele count and total number of alleles in the data.
```{r}
allele_freqs <- lapply(1:10, function(x){
  table(c(dbExample[[x*2]], dbExample[[1+x*2]]))/(2*nrow(dbExample))
})
names(allele_freqs) <- sub("\\.1", "", names(dbExample)[(1:10)*2])
head(allele_freqs)
```

## Single locus

The total number of alleles observed for $m$ contributors depends on
the number of loci, $L$, through the locus specific allele counts,
$N_l$, by $N(m) = \sum_{l=1}^L N_l(m)$. Hence, we can focus on
computing the distribution of $N_l(m)$ below.

The number of alleles at locus $l$, $N_l(m)$, follows a distribution,
where the number of alleles range from 1 (all $m$ profiles homozygous
with the same allele) to $2m$ (all $m$ profiles heterozygous sharing
no allele). For each value that $N_l(m)$ can attain there will
typically be several ways $m$ profiles can produce $N_l(m)$ distinct
alleles. For a given locus $l$, we use the function `Pnm_locus()` to 
compute $P(N_l(m) = n)$, for $n = 1,\dots,2m$, where we need to specify $m$
together with the allele frequencies.

```{r}
m <- 3
P3_D16 <- Pnm_locus(m = m, theta = 0, alleleProbs = allele_freqs$D16S539)
setNames(P3_D16, 1:(2*m))
```

The `theta` argument refers to the $\theta$ coefficient used to account
for genetic dependence of alleles in the Balding-Nicholls population genetic
model. Typical values ranges from $0$ (independent alleles, no subpopulation 
structure) to $0.03$ (mild correlation, some subpopulation structure).

We can also compute the probabilities for several $m$. Here we let $m$ 
range between 1 and 6 contributors.
```{r}
ms <- 1:6
Ps_D16 <- matrix(NA, ncol = length(ms), nrow = max(ms)*2, 
                 dimnames = list(alleles = 1:(max(ms)*2), noContrib=ms))
for(m in seq_along(ms)){
  Ps_D16[1:(ms[m]*2),m] <- Pnm_locus(m = ms[m], theta = 0, alleleProbs = allele_freqs$D16S539)
}
knitr::kable(Ps_D16, row.names = TRUE)
```

If we plot the outcome we see that for $m \in \{4, 5, 6\}$ the most probable
allele count is 4 indicating that D16S539 is not very polymorphic in this population.

```{r, fig.width=7, fig.asp=0.62}
par(mar = c(4,5,0,0))
plot(c(1,max(ms)*2), range(Ps_D16, na.rm = TRUE), type = "n",
     xlab = "no. of allele", ylab = "Probability")
for(m in seq_along(ms)) lines(1:(2*ms[m]), Ps_D16[1:(ms[m]*2),m], col = m)
legend("topright", bty = "n", title = "m", legend = ms, col = seq_along(ms), lty = 1)
```

The probability of masking can be calculated from the table above. Hence,
we need to compute $P(N_l(m) \le 2(m-1))$, which is done below.


```{r}
Ps_D16_cumsum <- apply(Ps_D16, 2, cumsum)
knitr::kable(Ps_D16_cumsum, row.names = TRUE)
```

We see that the probability that an $m$ person mixture is misconceived as 
an $m-1$ person mixture is 
```{r}
Pm_D16 <- setNames(rep(NA, max(ms)-1), paste0("P(N[",2:max(ms),"]<=",(1:(max(ms)-1))*2,")"))
for(m in 2:max(ms)) Pm_D16[m-1] <- Ps_D16_cumsum[2*(m-1),m]
Pm_D16
```
Hence, there is a probability of `r Pm_D16[2]` that a three-person mixture 
has at most four alleles, which implies is can be interpreted as a two-person mixture.

If we are interested in computing the locuswise probabilities for all loci
in our dataset, we simply use the `Pnm_all` function with `locuswise = TRUE`.

```{r}
knitr::kable(Pnm_all(m = 3, theta = 0, probs = allele_freqs, locuswise = TRUE), row.names = TRUE)
```

### Accumulating over loci

Since the loci are assumed independent, we can also compute the probability
that we see at most $n_0$ alleles at *all* markers:
$$
P(N_1(m) \le n_0, \dots, N_L(m)\le n_0) = \prod_{l=1}^L P(N_l(m) \le n_0)
$$

For instance, we may compute for $m\in\{1,\dots,6\}$ the probability of seeing 
at most $n_0 \in\{1,\dots,2m\}$ alleles for the ten STR markers in `allele_freqs`.

```{r}
locus_Ps <- lapply(ms, Pnm_all, theta = 0, probs = allele_freqs, locuswise = TRUE)
locus_Ps_cumsum <- lapply(locus_Ps, apply, 1, cumsum)
all_Ps_cumsum <- lapply(locus_Ps_cumsum, apply, 1, prod)
all_Ps_table <- matrix(NA, ncol = length(ms), nrow = max(ms)*2, 
                       dimnames = list(n0 = 1:(max(ms)*2), m = ms))
for(m in seq_along(ms)) all_Ps_table[1:(ms[m]*2), m] <- all_Ps_cumsum[[m]]
knitr::kable(all_Ps_table, row.names = TRUE)
```

Again we may extract the interesting entries the table as for a single locus.
```{r}
Pm_all <- setNames(rep(NA, max(ms)-1), paste0("P(N[",2:max(ms),"]<=",(1:(max(ms)-1))*2,")"))
for(m in 2:max(ms)) Pm_all[m-1] <- all_Ps_table[2*(m-1),m]
round(Pm_all, 4)
```

### $\theta$-correction

We can also look at the effect of the $\theta$-correction on the distribution
of observed number of alleles. The stronger the subpopulation effect (measured 
by increasing $\theta$), the fewer alleles will be observed.

```{r, fig.width=7, fig.asp=0.62}
par(mar = c(4,5,0,0))
P3_D16_t0.03 <- Pnm_locus(m = 3, theta = 0.03, alleleProbs = allele_freqs$D16S539)
P3_D16_t0.1 <- Pnm_locus(m = 3, theta = 0.1, alleleProbs = allele_freqs$D16S539)
plot(P3_D16_t0.03, type = "o", pch = 16, ylab = "Probability", xlab = "no. of alleles", col = 2)
points(P3_D16, type = "o", lty = 2, col = 1)
points(P3_D16_t0.1, type = "o", lty = 3, pch = 2, col = 3)
legend("topleft", bty = "n", title = expression(theta), col = 1:3,
       legend = c("0.00", "0.03", "0.10"), lty = c(2, 1, 3), pch = c(1, 16, 2))
```

## Summing over loci

In order to obtain $n$ alleles for $m$ contributors and $L$ loci, the
locus counts must sum to $n$. By convolution, we obtain
\begin{displaymath}
  P(N^{l{+}1}(m) = n) = \sum_{i=1}^I P(N^{l}(m) =
  n-i)P(N_{l+1}(m)=i),
\end{displaymath}
where $I = \min(2m,n{-}1)$ with $P(N_{l+1}(m)=i)$ denoting the
probability that locus $l{+}1$ has $i$ observed alleles, and
$P(N^l(m)=n)$ is the probability to observe $n$ alleles for $l$ loci
and $m$ profiles. 

We compute this by setting `locuswise = FALSE` in `Pnm_all`.
```{r, fig.width=7, fig.asp=0.62}
par(mar = c(4,5,0,0))
P3_all <- Pnm_all(m = 3, theta = 0, probs = allele_freqs, locuswise = FALSE)
plot(P3_all, xlab = "no. of alleles", ylab = "Probability", type = "o")
```

## Posterior probability for the number of contributors

Assume that we specify the prior, $P(m)$, on $m$ based on some general idea on the 
number of contributors we typically see, or based on other background information
regarding the case. Then $P(m)$ reflects the belief we have about $m$ prior to 
seeing the DNA mixture.

Using Bayes' theorem, we can express the probability distribution for
the number of contributors, $m$, given the number of observed alleles,
$n_{l}$:
$$
  P(m\mid n_{l}) = \frac{P(n_{l}\mid m)P(m)}{P(n_{l})} =
  \frac{P(N_l(m) = n_{l})P(m)}{\sum_m P(N_l(m) = n_{l})P(m)},
$$
where $P(m)$ is a prior distribution on the number of contributors,
$m$. We may also generalise this to including all information, 
$\underline{n} = (n_1, \dots, n_L)$, from 
the loci based on the independent assumption.

$$
  P(m\mid \underline{n}) = \frac{P(\underline{n}\mid m)P(m)}{P(\underline{n})} =
  \frac{P(m)\prod_{l=1}^L P(N_l(m) = n_{l})}{\sum_m\left\{ P(m)\prod_{l=1}^LP(N_l(m) = n_{l})\right\}},
$$

Assume that $P(m)$ is given by 
```{r, fig.width=7, fig.asp=0.62}
par(mar = c(4,5,0,0))
m_prior <- c(0.2,0.35,0.3,0.2,0.1,0.025)
m_prior <- m_prior/sum(m_prior)
barplot(m_prior, names.arg = seq_along(m_prior), ylab = "Prior probability,"~italic(P(m)), xlab = expression(italic(m)))
```

Let $n_{vWA} = 4$. Then we can update the belief based on this information using 
the expression above.
```{r}
n_vWA <- 4
P_vWA <- lapply(ms, Pnm_locus, theta = 0, alleleProbs = allele_freqs$vWA)
P_vWA_n <- sapply(P_vWA, function(p) ifelse(length(p) < n_vWA, 0, p[n_vWA]))
Pn_vWA <- (m_prior * P_vWA_n)/sum(m_prior * P_vWA_n)
```

Hence, the change in belief about $m$ can be seen below, where the first row
is the prior $P(m)$, the second $P(N_{vWA}(m) = n_{vWA})$ and the last 
$P(n \mid N_{vWA}(m) = n_{vWA})$:

```{r}
knitr::kable(rbind(m_prior, P_vWA_n, Pn_vWA), col.names = paste(ms))
```

We see that $m =$ `r which.max(Pn_vWA)` is the most probable after seeing 
`r n_vWA` alleles at vWA.

Assuming that we observe $n_l = 4$ for all loci, $l=1,\dots,L$, we get
```{r}
get_pn <- function(p, n){
  pn <- split(as.data.frame(p), n)
  p_n <- unlist(lapply(names(pn), function(n_l) if(ncol(pn[[n_l]])<n_l) rep(0,nrow(pn[[n_l]])) else pn[[n_l]][,n_l]))
  prod(p_n)
}

n_loci <- rep(4, length(allele_freqs))
P_all <- lapply(ms, Pnm_all, theta = 0, probs = allele_freqs, locuswise = TRUE)
P_all_n <- sapply(P_all, get_pn, n_loci)
Pn_all <- (m_prior * P_all_n)/sum(m_prior * P_all_n)
```

Hence, the posterior probability of $m$ after observing four alleles at all loci
is given by 
```{r}
setNames(Pn_all, paste0("P(m=", seq_along(Pn_all),"|n=4)"))
```
